---
title: "07_mediation_analysis"
author: "Jagadeesh Puvvula"
date: "2024-12-29"
output: pdf_document
---

```{r}
all_results<- read_csv("~/Documents/air_polln_rna_seq/results/pm_bc_twas_dec_29.csv") |>
  filter(FDR < 0.05) |>
  filter(str_starts(predictor_variable, "bc")) |>
  distinct(gene, gene_symbol)
```

#Clean RNAseq, exposure, covariate data
```{r}
load("~/Documents/air_polln_rna_seq/crib_data/for_mediation_analy.rda")

#filtered genes that are significantly associated with air pollution
rna_seq_med <- edge_df |>
  filter(Gene %in% all_results$gene) |>
  t()

# Set the first row as column names and remove it from the data
colnames(rna_seq_med) <- rna_seq_med[1, ]
rna_seq_med <- rna_seq_med[-1, ] 

#scaling rna seq values
rna_seq_med_numeric <- matrix(as.numeric(rna_seq_med), nrow=nrow(rna_seq_med), ncol=ncol(rna_seq_med))
rownames(rna_seq_med_numeric) <- rownames(rna_seq_med)
colnames(rna_seq_med_numeric) <- colnames(rna_seq_med)
scaled_rna_seq_med <- scale(rna_seq_med_numeric)

```

#mediation analysis
#alpha_hat: coefficient estimates of exposure (X) –> mediators (M) (adjusted for covariates).
#alpha_se: standard error for alpha.
#beta_hat: coefficient estimates of mediators (M) –> outcome (Y) (adjusted for covariates and exposure).
#beta_se: standard error for beta.
#IDE: mediation (indirect) effect, i.e., alpha*beta.
#rimp: relative importance of the mediator.
#pmax: joint raw p-value of selected significant mediator (based on divide-aggregate compositenull test [DACT] method).
```{r}
classicHima.fit <- classicHIMA(
  X = as.numeric(df_analysis$pm_26),
  Y = as.factor(df_analysis$ptb),
  M = scaled_rna_seq_med,
  #COV.XM = df_analysis[, c("ga_at_delivery", "maternal_age")],
  #COV.MY = df_analysis[, c("ga_at_delivery", "maternal_age")],
  penalty = c("lasso"),
  Bonfcut = 0.1,
  Y.type = "binary",
  scale = T, 
  verbose = TRUE,
  parallel = T,
  ncore = 10
  )
```

#automating
```{r}
# Initialize empty dataframe
final_results <- data.frame()

for (i in 1:141) {
  pm_col <- paste0("bc_", i)
  
  tryCatch({
    print(paste("Processing:", pm_col))
    
    result <- classicHIMA(
      X = as.numeric(df_analysis[[pm_col]]),
      Y = as.factor(df_analysis$ptb),
      M = scaled_rna_seq_med,
      penalty = c("lasso"),
      Y.type = "binary",
      Bonfcut = 0.05,
      scale = TRUE,
      verbose = TRUE,
      parallel = TRUE,
      ncore = 10
    )
    
    print("Result structure:")
    print(str(result))
    
    # If the result is not null and contains data, process it
    if (!is.null(result) && nrow(result) > 0) {
      print("Full result:")
      print(result)
      
      # Convert result to dataframe
      result_df <- as.data.frame(result)
      
      # Add the pm_variable column
      result_df$pm_variable <- pm_col
      
      # Append to final_results
      if (nrow(final_results) == 0) {
        final_results <- result_df
      } else {
        final_results <- rbind(final_results, result_df)
      }
      
      print("Current final_results dimensions:")
      print(dim(final_results))
    } else {
      print(paste("No valid data for", pm_col))
    }
    
    print(paste("Completed analysis for", pm_col))
    
  }, error = function(e) {
    message(paste("Error in", pm_col, ":", e$message))
  })
}
```

```{r}
write_csv(final_results, "~/Documents/air_polln_rna_seq/results/mediation_bc.csv")
```


